---
title: "Student marks analysis"
author: "Jarun ~ Mohana ~ Aravind"
output:
  pdf_document: default
  word_document: default
  html_document: default
---



Importing all the datasets
```{r}
library(tidyverse) # metapackage with lots of helpful functions
library(dplyr)
#install.packages("reshape2")
library(reshape2)
```

Importing the dataset
```{r}

data = read.csv("studentdata.csv")
str(data)

```

```{r}
print(unique(data$gender))
print(unique(data$race.ethnicity))
print(unique(data$parental.level.of.education))
print(unique(data$lunch))
print(unique(data$test.preparation.course))
```


```{r}
head(data)
```


General Description

Affectors:
1) Gender
2) Race ethnicity
3) Parental level of education
4) Lunch
5) Test preparation course

Marks:
1) Math score
2) Reading score
3) Writing score



# Analysis on Race ethinicity over students' marks


```{r}
data %>%
    group_by(race.ethnicity) %>%
    summarise(mathMean = mean(math.score),readingMean =  mean(reading.score),writingMean = mean(writing.score)) %>% 
  melt(id.vars="race.ethnicity", variable.name = "Marks") %>% 
  ggplot(aes(x=Marks, y=value, fill=race.ethnicity)) +
  geom_bar(stat="identity", position = position_dodge())

```


#### Result
In general there are no large difference at all. Maybe group E is slighty better than others 
but generally, all the groups are fine in the total pool.



```{r}

mathpass = data %>% 
  group_by(race.ethnicity) %>% 
  filter(math.score>40) %>% 
  count(race.ethnicity)

writingpass = data %>% 
  group_by(race.ethnicity) %>% 
  filter(writing.score>40) %>% 
  count(race.ethnicity)

readingpass = data %>% 
  group_by(race.ethnicity) %>% 
  filter(reading.score>40) %>% 
  count(race.ethnicity)

total = data %>% 
  group_by(race.ethnicity) %>% 
  count(race.ethnicity)

total$mathPassPercent = mathpass$n / total$n * 100
total$readingPassPercent = readingpass$n / total$n * 100
total$writingPassPercent = writingpass$n / total$n * 100

total = subset(total, select = -c(n))

melt(total, id.vars="race.ethnicity", variable.name = "passPercent") %>% 
  ggplot(aes(x=passPercent, y=value, fill=race.ethnicity)) +
  geom_bar(stat="identity", position = position_dodge())

```

#### Result
In general there are no large difference at all. All the groups have
good pass percentage.


```{r}
math = data %>% 
  group_by(race.ethnicity) %>% 
  filter(math.score>80) %>% 
  count(race.ethnicity)

writing = data %>% 
  group_by(race.ethnicity) %>% 
  filter(writing.score>80) %>% 
  count(race.ethnicity)

reading = data %>% 
  group_by(race.ethnicity) %>% 
  filter(reading.score>80) %>% 
  count(race.ethnicity)

total = data %>% 
  group_by(race.ethnicity) %>% 
  count(race.ethnicity)

total$mathAPercent = math$n / total$n * 100
total$readingAPercent = reading$n / total$n * 100
total$writingAPercent = writing$n / total$n * 100

total = subset(total, select = -c(n))

melt(total, id.vars="race.ethnicity", variable.name = "GoodMarkPercent") %>% 
  ggplot(aes(x=GoodMarkPercent, y=value, fill=race.ethnicity)) +
  geom_bar(stat="identity", position = position_dodge())
```

#### Result
We see a huge difference here. A large chunk of students of Group E are 
talented in math and reading. 


```{r}

df = data %>% 
  filter(race.ethnicity == "group E") %>% 
  select(race.ethnicity, math.score, writing.score, reading.score)  

hist(x= df$math.score, main = "Histogram of math scores of Group E")

```

```{r}
df = data %>% 
  filter(race.ethnicity == "group A") %>% 
  select(race.ethnicity, math.score, writing.score, reading.score)  

hist(x= df$math.score, main = "Histogram of math scores of Group A")

```


#### Result
Group A is not doing good with math.



## Gender

```{r}
data %>%
    group_by(gender) %>%
    summarise(mathMean = mean(math.score),readingMean =  mean(reading.score),writingMean = mean(writing.score)) %>% 
  melt(id.vars="gender", variable.name = "Marks") %>% 
  ggplot(aes(x=Marks, y=value, fill=gender)) +
  geom_bar(stat="identity", position = position_dodge())
```

#### Result
In general there are no large difference found.


```{r}

mathpass = data %>% 
  group_by(gender) %>% 
  filter(math.score>40) %>% 
  count(gender)

writingpass = data %>% 
  group_by(gender) %>% 
  filter(writing.score>40) %>% 
  count(gender)

readingpass = data %>% 
  group_by(gender) %>% 
  filter(reading.score>40) %>% 
  count(gender)

total = data %>% 
  group_by(gender) %>% 
  count(gender)

total$mathPassPercent = mathpass$n / total$n * 100
total$readingPassPercent = readingpass$n / total$n * 100
total$writingPassPercent = writingpass$n / total$n * 100

total = subset(total, select = -c(n))

melt(total, id.vars="gender", variable.name = "passPercent") %>% 
  ggplot(aes(x=passPercent, y=value, fill=gender)) +
  geom_bar(stat="identity", position = position_dodge())

```

#### Result
Both boys and girls have good pass percentage over all the subjects.
Boys have a little edge over math and girls have it over reading and writing.


```{r}

math = data %>% 
  group_by(gender) %>% 
  filter(math.score>80) %>% 
  count(gender)

writing = data %>% 
  group_by(gender) %>% 
  filter(writing.score>80) %>% 
  count(gender)

reading = data %>% 
  group_by(gender) %>% 
  filter(reading.score>80) %>% 
  count(gender)

total = data %>% 
  group_by(gender) %>% 
  count(gender)

total$mathAPercent = math$n / total$n * 100
total$readingAPercent = reading$n / total$n * 100
total$writingAPercent = writing$n / total$n * 100

total = subset(total, select = -c(n))

melt(total, id.vars="gender", variable.name = "GoodMarkPercent") %>% 
  ggplot(aes(x=GoodMarkPercent, y=value, fill=gender)) +
  geom_bar(stat="identity", position = position_dodge())

```

#### Result
Adding to the inference we got in the last plot, we find boys perform well in math
and girls perform well in reading and writing.



Number of students who has scored above 80 in all subjects
grouped by gender.

```{r}

data %>% 
    group_by(gender) %>%
    select(math.score,reading.score,writing.score) %>%
    filter(math.score > 80, reading.score > 80, writing.score > 80) %>% 
    count(gender) %>% 
    ggplot(data = ., aes(x = gender, y = n, 
    fill = gender)) + geom_bar(stat = "identity")

```
#### Result
Overall, Female students are more studious than male students.


## Level of Education Differences

```{r}
data %>%
    group_by(parental.level.of.education) %>%
    summarise(mathMean = mean(math.score),readingMean =  mean(reading.score),writingMean = mean(writing.score)) %>% 
  melt(id.vars="parental.level.of.education", variable.name = "Marks") %>% 
  ggplot(aes(x=Marks, y=value, fill=parental.level.of.education)) +
  geom_bar(stat="identity", position = position_dodge())
```

```{r}

mathpass = data %>% 
  group_by(parental.level.of.education) %>% 
  filter(math.score>40) %>% 
  count(parental.level.of.education)

writingpass = data %>% 
  group_by(parental.level.of.education) %>% 
  filter(writing.score>40) %>% 
  count(parental.level.of.education)

readingpass = data %>% 
  group_by(parental.level.of.education) %>% 
  filter(reading.score>40) %>% 
  count(parental.level.of.education)

total = data %>% 
  group_by(parental.level.of.education) %>% 
  count(parental.level.of.education)

total$mathPassPercent = mathpass$n / total$n * 100
total$readingPassPercent = readingpass$n / total$n * 100
total$writingPassPercent = writingpass$n / total$n * 100

total = subset(total, select = -c(n))

melt(total, id.vars="parental.level.of.education", variable.name = "passPercent") %>% 
  ggplot(aes(x=passPercent, y=value, fill=parental.level.of.education)) +
  geom_bar(stat="identity", position = position_dodge())

```


```{r}

math = data %>% 
  group_by(parental.level.of.education) %>% 
  filter(math.score>80) %>% 
  count(parental.level.of.education)

writing = data %>% 
  group_by(parental.level.of.education) %>% 
  filter(writing.score>80) %>% 
  count(parental.level.of.education)

reading = data %>% 
  group_by(parental.level.of.education) %>% 
  filter(reading.score>80) %>% 
  count(parental.level.of.education)

total = data %>% 
  group_by(parental.level.of.education) %>% 
  count(parental.level.of.education)

total$mathAPercent = math$n / total$n * 100
total$readingAPercent = reading$n / total$n * 100
total$writingAPercent = writing$n / total$n * 100

total = subset(total, select = -c(n))

melt(total, id.vars="parental.level.of.education", variable.name = "GoodMarkPercent") %>% 
  ggplot(aes(x=GoodMarkPercent, y=value, fill=parental.level.of.education)) +
  geom_bar(stat="identity", position = position_dodge())

```



```{r}

data %>% 
    group_by(parental.level.of.education) %>%
    select(math.score,reading.score,writing.score) %>%
    filter(math.score > 80, reading.score > 80, writing.score > 80) %>% 
    count(parental.level.of.education) %>% 
    ggplot(data = ., aes(x = parental.level.of.education, y = n, 
    fill = parental.level.of.education)) + geom_bar(stat = "identity")

```

Result
Generally, the results are the same. Therefore, the scores does not changes with educational level. 




## Test Preperation Course Effect on Scores


```{r}
data %>% 
    group_by(test.preparation.course) %>%
    summarise(mathMean = mean(math.score),readingMean =  mean(reading.score),writingMean = mean(writing.score)) %>%
    mutate(totalScores = mathMean + readingMean + writingMean) %>%
    ggplot(data = ., aes(x = test.preparation.course, y = totalScores, 
    fill = test.preparation.course)) + geom_bar(stat = "identity") + 
    labs(title="Test Preparation Comparison",
    subtitle="Total Scores") +
    theme(axis.text.x = element_text(angle=65, vjust=0.6))
```

Result
Total scores of competeled ones is higher than who had not taken the preparation course.


## Analyzing Lunch effects on Scores

```{r}
totalLunch <- table(data$lunch)
totalLunch
```
```{r}
lunch <- data %>% 
    group_by(lunch) %>%
    summarise(mathTotal = sum(math.score),readingTotal =  sum(reading.score),writingTotal = sum(writing.score)) %>%
    mutate(totalScores = mathTotal + readingTotal + writingTotal) 
lunch %>%
    ggplot(data = ., aes(x = lunch, y = totalScores, 
    fill = lunch)) + geom_bar(stat = "identity") + 
    labs(title="Lunch Comparison",
    subtitle="Total Scores") +
    theme(axis.text.x = element_text(angle=65, vjust=0.6))
```

```{r}


mathpass = data %>% 
  group_by(lunch) %>% 
  filter(math.score>40) %>% 
  count(lunch)

writingpass = data %>% 
  group_by(lunch) %>% 
  filter(writing.score>40) %>% 
  count(lunch)

readingpass = data %>% 
  group_by(lunch) %>% 
  filter(reading.score>40) %>% 
  count(lunch)

total = data %>% 
  group_by(lunch) %>% 
  count(lunch)

total$mathPassPercent = mathpass$n / total$n * 100
total$readingPassPercent = readingpass$n / total$n * 100
total$writingPassPercent = writingpass$n / total$n * 100

total = subset(total, select = -c(n))

melt(total, id.vars="lunch", variable.name = "passPercent") %>% 
  ggplot(aes(x=passPercent, y=value, fill=lunch)) +
  geom_bar(stat="identity", position = position_dodge())
```

```{r}

math = data %>% 
  group_by(lunch) %>% 
  filter(math.score>80) %>% 
  count(lunch)

writing = data %>% 
  group_by(lunch) %>% 
  filter(writing.score>80) %>% 
  count(lunch)

reading = data %>% 
  group_by(lunch) %>% 
  filter(reading.score>80) %>% 
  count(lunch)

total = data %>% 
  group_by(lunch) %>% 
  count(lunch)

total$mathAPercent = math$n / total$n * 100
total$readingAPercent = reading$n / total$n * 100
total$writingAPercent = writing$n / total$n * 100

total = subset(total, select = -c(n))

melt(total, id.vars="lunch", variable.name = "GoodMarkPercent") %>% 
  ggplot(aes(x=GoodMarkPercent, y=value, fill=lunch)) +
  geom_bar(stat="identity", position = position_dodge())

```


Result
In general, standard lunches seen more effective on the students.

## Correlation of Scores

```{r}
library(corrplot)
library(Hmisc)
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}
res2<-rcorr(as.matrix(data[,6:8]))
flattenCorrMatrix(res2$r, res2$P)

```

# Conclusion
There is highly correlation between scores. Thus, the students who takes low scores at one of area could take another low score and high scores take another high scores. 


Being female, parents having a master degree and being a group E ethnicity is a advantage in education or maybe for carrier.





# Machine learning

```{r}
library(superml)
library(caTools)
library(caret)
library(quantreg)
library(data.table)
library(reshape2)
```


```{r}
df = data
```

```{r}

split = sample.split(df, SplitRatio = 0.9)
train = subset(df, split=="TRUE")
test = subset(df, split=="FALSE")
head(train)
head(test)
```

Train set size: 875
Test set size: 125



### Label Encoding

```{r}

data = read.csv("studentdata.csv")

newdata <- tocategorical( data)
newdata


```


```{r}
genderlabel = LabelEncoder$new()
train$gender = genderlabel$fit_transform(train$gender)

racelabel = LabelEncoder$new()
train$race.ethnicity = racelabel$fit_transform(train$race.ethnicity)

parentallabel <- LabelEncoder$new()
train$parental.level.of.education = parentallabel$fit_transform(train$parental.level.of.education)

lunchlabel <- LabelEncoder$new()
train$lunch = lunchlabel$fit_transform(train$lunch)

testlabel <- LabelEncoder$new()
train$test.preparation.course = testlabel$fit_transform(train$test.preparation.course)

head(train)
```



Train test split:

```{r}

train_x = subset(train, select = c(gender, race.ethnicity, parental.level.of.education, lunch, test.preparation.course))
train_y = subset(train, select = -c(gender, race.ethnicity, parental.level.of.education, lunch, test.preparation.course))


test_x = subset(test, select = c(gender, race.ethnicity, parental.level.of.education, lunch, test.preparation.course))
test_y = subset(test, select = -c(gender, race.ethnicity, parental.level.of.education, lunch, test.preparation.course))


head(train_x)
head(train_y)
head(test_x)
head(test_y)

```


```{r}

test_x$gender = genderlabel$transform(test_x$gender)
test_x$race.ethnicity = racelabel$transform(test_x$race.ethnicity)
test_x$parental.level.of.education = parentallabel$transform(test_x$parental.level.of.education)
test_x$lunch = lunchlabel$transform(test_x$lunch)
test_x$test.preparation.course = testlabel$transform(test_x$test.preparation.course)

head(test_x)
```




### Linear model between math score and all factors

```{r}
lm_model_math = lm(math.score ~ gender+race.ethnicity+parental.level.of.education+lunch+test.preparation.course, train)
summary(lm_model_math)
view(data)
```


```{r}
lm_pred_math = predict(lm_model_math, test_x)
lm_rmse_math  = sqrt(sum((lm_pred_math- test_y$math.score)^2)/125)
lm_rmse_math
```


```{r}
i = seq(1:length(test_y$math.score))
lm_table = data.frame(i, lm_pred_math, test_y$math.score)
head(lm_table)
```

```{r}
ggplot(data=lm_table) + geom_point(aes(x=i,y=lm_pred_math, color = "blue")) + 
  geom_point(aes(x=i,y=test_y.math.score, color = "red")) 
```



### Linear model between writing score and all factors

```{r}
lm_model_writing = lm(writing.score ~ gender+race.ethnicity+parental.level.of.education+lunch+test.preparation.course, train)
summary(lm_model_writing)

```


```{r}
lm_pred_writing = predict(lm_model_writing, test_x)
lm_rmse_writing = sqrt(sum((lm_pred_writing- test_y$writing.score)^2)/125)
lm_rmse_writing
```


### Linear model between reading score and all factors

```{r}
lm_model_reading = lm(reading.score ~ gender+race.ethnicity+parental.level.of.education+lunch+test.preparation.course, train)
summary(lm_model_reading)

```


```{r}
lm_pred_reading = predict(lm_model_reading, test_x)
lm_rmse_reading = sqrt(sum((lm_pred_reading- test_y$reading.score)^2)/125)
lm_rmse_reading
```


## Anova

```{r}
aov_model_math = aov(math.score ~ gender+race.ethnicity+parental.level.of.education+lunch+test.preparation.course, train)
aov_pred_math = predict(aov_model_math, test_x)
aov_rmse_math = sqrt(sum((aov_pred_math- test_y$math.score)^2)/125)
aov_rmse_math
```

```{r}
aov_model_writing = aov(writing.score ~ gender+race.ethnicity+parental.level.of.education+lunch+test.preparation.course, train)
aov_pred_writing = predict(aov_model_writing, test_x)
aov_rmse_writing = sqrt(sum((aov_pred_writing- test_y$writing.score)^2)/125)
aov_rmse_writing
```

```{r}
aov_model_reading = aov(reading.score ~ gender+race.ethnicity+parental.level.of.education+lunch+test.preparation.course, train)
aov_pred_reading = predict(aov_model_reading, test_x)
aov_rmse_reading = sqrt(sum((aov_pred_reading- test_y$reading.score)^2)/125)
aov_rmse_reading
```


```{r}

i = seq(1:length(test_y$math.score))
aov_table = data.frame(i, aov_pred_math, test_y$math.score)

ggplot(data=aov_table) + geom_point(aes(x=i,y=aov_pred_math, color = "blue")) + 
  geom_point(aes(x=i,y=test_y.math.score, color = "red")) 

```



## Random forest regression

```{r}
library(caTools)
library(randomForest)
library(Metrics)
```


### Random forest on Math score

```{r}


set.seed(123)


regressor = randomForest(x = train_x,
                         y = train_y$math.score,
                         ntree = 500)

y_pred = predict(regressor, test_x)


i = seq(1:length(test_y$math.score))
regression_table = data.frame(i, y_pred, test_y$math.score)
head(regression_table)


```

```{r}
rf_rmse_math =  rmse(y_pred, test_y$math.score)
rf_rmse_math
```


```{r}
ggplot(data=regression_table) + geom_point(aes(x=i,y=y_pred, color = "blue")) + 
  geom_point(aes(x=i,y=test_y.math.score, color = "red")) 

```


### Random forest on writing score

```{r}

regressor = randomForest(x = train_x,
                         y = train_y$writing.score,
                         ntree = 500)

y_pred = predict(regressor, test_x)




```

Sum of squared error:
```{r}
rf_rmse_writing = rmse(y_pred, test_y$writing.score)
rf_rmse_writing
```




### Random forest on reading score

```{r}

regressor = randomForest(x = train_x,
                         y = train_y$reading.score,
                         ntree = 500)

y_pred = predict(regressor, test_x)



```


```{r}
rf_rmse_reading = rmse(y_pred, test_y$reading.score)
rf_rmse_reading
```



## Analysing the models

```{r}
name = c("Math RMSE", "Reading RMSE", "Writing RMSE")
reg_rmse = c(lm_rmse_math, lm_rmse_reading, lm_rmse_writing)
aov_rmse = c(aov_rmse_math, aov_rmse_reading, aov_rmse_writing)
rf_rmse = c(rf_rmse_math, rf_rmse_reading, rf_rmse_writing)

data.frame(name, reg_rmse, aov_rmse, rf_rmse)
```








